---
title: 'STA305H5S: Project'
author: "Jack Chamoun Barbar, Mir Asim Ali, Parth Tyagi"
date: "April 5, 2026"
output:
  pdf_document:
    number_sections: true
  word_document: default
spacing: double
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- "\\fancyhead[L]{STA305H5-Experimental Design}"
- \fancyfoot[CO,CE]{}
- \fancyfoot[LE,RO]{\thepage}
- \usepackage{setspace}\doublespacing
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

# The libraries needed for our project
library(scidesignR)
library(ggplot2)
library(dplyr)
library(tidyr)
library(car)
library(coin)
```

\newpage 
\tableofcontents 
\newpage

# Part 1: Comparing Two Treatments

## Study Design Definitions

* **Research Question:** Is there a statistically significant difference in the mean wheat yield between Treatment A and Treatment B? In other words, we are testing: $H_0:\mu_A=\mu_B$ vs $H_1:\mu_A \neq \mu_B$, where
$\mu_A$ and $\mu_B$ refer to the mean wheat yields of treatments A and B, respectively.
* **Experimental Units:** The wheat plots, which are randomly assigned to receive the treatments.
* **Response Variable:** Wheat yield. It is represented as `fert` in the `fertdat`
dataset. Note that wheat yield is a quantitative continuous response variable
(as the data are collected by measuring and expressed on a continuous scale).
* **Treatment Factor:** Level/type of fertilizer, which is expressed as `trt` the
`fertdat` dataset, with two levels (A and B).

## Exploratory Data Analysis (EDA)
Exploratory Data Analysis (EDA) is the initial phase of data science that uses descriptive statistics and visual techniques  to uncover patterns, identify anomalies, and understand the underlying structure of a dataset. It is essential because it prevents flawed modeling by informing data cleaning processes, testing necessary statistical assumptions, and guiding the selection of appropriate statistical tests.

\large
```{r eda, fig.width=4, fig.height=3.5, fig.align='center'}
# Load the dataset
data("fertdat", package = "scidesignR")


summary_stats <- fertdat %>%
  group_by(trt) %>%
  summarize(
    n = n(),
    Mean = round(mean(fert, na.rm = TRUE), 4),
    SD = round(sd(fert, na.rm = TRUE), 4),
    Min = min(fert, na.rm = TRUE),
    Max = max(fert, na.rm = TRUE)
  )
knitr::kable(summary_stats, caption = "Summary Statistics of Wheat Yield by Fertilizer")

# A boxplot
ggplot(fertdat, aes(x = trt, y = fert, fill = trt)) +
  geom_boxplot(alpha = 0.7) +
  theme_minimal() +
  labs(title = "Wheat Yield Distribution by Fertilizer Type",
       x = "Type of Fertilizer",
       y = "Wheat Yield") +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5))
```
\normalsize

## Checking Assumptions

To proceed with a standard two-sample t-test, we must check for **normality** of the residuals
and **homoscedasticity** of variance, i.e., constant variance.


```{r model_setup, include=FALSE}
mod <- lm(fert ~ trt, data = fertdat)
```

```{r qq_plot, fig.width=4.5, fig.height=3.8, fig.align='center', out.width='65%'}
qqnorm(resid(mod), main = "Q-Q Plot of Residuals"); qqline(resid(mod), col = "red")
```
\normalsize

\newpage

```{r residual_plot, fig.width=4.5, fig.height=4, fig.align='center'}
plot(fitted(mod), resid(mod),
     main = "Residuals vs Fitted",
     xlab = "Fitted Values",
     ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)
```
Since plots can be subjective, it is better to supplement with more formal/robust tests. Now, we use the Shapiro and Levene's tests, which assess for normality and equality of population variances
(i.e, $\sigma^2_A = \sigma^2_B$), respectively. We chose Levene's test as it is very robust against
non-normality, unlike Bartlett's test.

\vspace{-2em}
```{r Formal tests, comment = NA, strip.white = TRUE}
# Normality check
shapiro.test(resid(mod))

# Levene's test for equal variances
leveneTest(fert ~ trt, data = fertdat)
```

**Assumption Check Conclusion:** The assumption of normality is met, as the Q-Q plot points closely follow the theoretical line and the Shapiro-Wilk test yields a non-significant p-value of 0.8418. However, the assumption of homoscedasticity is violated because the Residuals vs Fitted plot displays an unequal spread and Levene’s test results in a significant p-value of 0.04837. Thus, we should proceed with a Welch's two-sample t-test instead of a standard two-sample t-test to account for the unequal variances ($\sigma^2_A \neq \sigma^2_B$).

## Hypothesis Testing

Let $\mu_A$ and $\mu_B$ be the true mean wheat yields for treatments A and B, respectively.

* **Null Hypothesis ($H_0$):** $\mu_A = \mu_B$
* **Alternative Hypothesis ($H_1$):** $\mu_A \neq \mu_B$

\vspace{-2.5em}
```{r Testing, comment = NA, strip.white = TRUE}
# Welch's two-sample t-test
welch_test <- t.test(fert ~ trt, data = fertdat, var.equal = FALSE)
welch_test
```

## Results and Interpretation

Based on the R output for Welch's two-sample t-test:

* **Test Statistic:** $t = `r round(welch_test$statistic, 4)`$
* **Degrees of Freedom:** $df = `r round(welch_test$parameter, 4)`$
* **P-value:** $`r format.pval(welch_test$p.value)`$
* **95% Confidence Interval:** $[`r round(welch_test$conf.int[1], 4)`, `r round(welch_test$conf.int[2], 4)`]$

**Conclusion:** With a large p-value of 0.6168, we fail to reject the null hypothesis, as there is insufficient evidence to conclude a significant difference in true mean wheat yields between fertilizers A and B. Furthermore, we are 95% confident that the true difference in mean yields ($\mu_A - \mu_B$) lies between -5.86 and 9.26, an interval that includes zero and is consistent with the lack of a statistically significant difference.


# Part 2: Power & Sample Size

## Practical Effect Size and Required Sample Size
We use the built-in `power.t.test()` function in R to compute the Effect Size
and the sample size required to achieve 80% power.

\vspace{-3em}
```{r effect-size, comment = NA, strip.white = TRUE}
delta_est <- 22.1833 - 20.4833

sd_pooled <- sigma(mod)

alpha_level <- 0.05; target_power <- 0.80

exact_calc <- power.t.test(delta = delta_est,
                           sd = sd_pooled,
                           sig.level = alpha_level,
                           power = target_power,
                           type = "two.sample",
                           alternative = "two.sided")

exact_calc
```

Based on the linear model from Part 1, the estimated (unstandardized) effect size ($\Delta = |\bar{x}_A - \bar{x}_B|$) is `r round(delta_est, 4)`, and the pooled standard deviation is $s_{pooled}\approx$ `r round(sd_pooled, 4)`.
Thus, our standardized effect size (Cohen's $d$) is $d = \frac{|\bar{x}_A - \bar{x}_B|}{s_{pooled}} = \frac{1.7}{5.6508} \approx 0.3008$. The required sample size to achieve 80% power at a significance level of $\alpha = 0.05$ is **175** (after rounding up to the next integer) **per group**.

## Simulated Power
This code calculates statistical power by simulating two data groups 10,000 times, running a t-test on each iteration, and finding the proportion of p-values that fall below the significance threshold $\alpha=0.05$.

```{r Simulated Power}
set.seed(123)  # So that whoever runs this gets same exact number
m <- 10000
p_values <- numeric(m) # To store the p-values in this (will use after the loop)

for(i in 1:m) {
  sim_A <- rnorm(175, mean = 0, sd = sd_pooled)
  sim_B <- rnorm(175, mean = delta_est, sd = sd_pooled)
  test_result <- t.test(sim_A, sim_B, var.equal = FALSE)
  p_values[i] <- test_result$p.value
}

simulated_power <- mean(p_values < alpha_level) # Proportion of times we reject
cat("Simulated Power:", simulated_power, "\n")
```

## Discussion on Changing Parameters

* **If the effect size is smaller:** The required sample size **increases**. A smaller true difference is harder to detect against random noise, so more data is needed to correctly reject the null hypothesis.
* **If the variability is larger:** The required sample size **increases**. A higher standard deviation creates more overlap between the two groups, requiring more observations to clearly isolate the treatment effect.

# Part 3: Comparing More Than Two Treatments

## 3A: One-Way ANOVA

### Study Design Definitions

* **Research Question:** Is there a statistically significant difference in mean pain scores among the three treatments (A, B, and C)? Formally, we test: $H_0: \mu_A = \mu_B = \mu_C$ vs $H_1:$ at least one mean differs, where $\mu_A$, $\mu_B$, and $\mu_C$ represent the true mean pain scores for treatments A, B, and C, respectively.
* **Experimental Units:** The individual subjects/patients who receive the treatments.
* **Response Variable:** Pain score (`pain`), measured on a scale from 0 to 10.
* **Treatment Factor:** Treatment type (`trt`), with three levels: A, B, and C.

### Exploratory Data Analysis

```{r eda_3a, fig.width=4.5, fig.height=3.5, fig.align='center'}
# Load the dataset
data("painstudy", package = "scidesignR")

# Summary statistics by treatment
summary_pain <- painstudy %>%
  group_by(trt) %>%
  summarize(
    n = n(),
    Mean = round(mean(pain), 4),
    SD = round(sd(pain), 4),
    Min = min(pain),
    Max = max(pain)
  )
knitr::kable(summary_pain, caption = "Summary Statistics of Pain Scores by Treatment")

# Boxplot
ggplot(painstudy, aes(x = trt, y = pain, fill = trt)) +
  geom_boxplot(alpha = 0.7) +
  theme_minimal() +
  labs(title = "Pain Score Distribution by Treatment",
       x = "Treatment",
       y = "Pain Score") +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5))
```

From the boxplot and summary table, we can see noticeable differences between the three treatment groups. Treatment C has the lowest mean pain score at 3.64, Treatment A is in the middle with 6.20, and Treatment B has the highest at 7.32. The spread of the data looks fairly similar across groups.

### Model Fitting and Assumption Checking

We fit a one-way ANOVA model to compare pain scores across the three treatments.

```{r anova_model_3a, include=FALSE}
anova_mod <- aov(pain ~ trt, data = painstudy)
```

```{r qq_3a, fig.width=4.5, fig.height=3.8, fig.align='center', out.width='65%'}
qqnorm(resid(anova_mod), main = "Q-Q Plot of Residuals"); qqline(resid(anova_mod), col = "red")
```

```{r resid_3a, fig.width=4.5, fig.height=3.8, fig.align='center', out.width='65%'}
plot(fitted(anova_mod), resid(anova_mod),
     main = "Residuals vs Fitted",
     xlab = "Fitted Values",
     ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)
```

```{r assumption_tests_3a, comment = NA, strip.white = TRUE}
# Shapiro-Wilk test for normality
shapiro.test(resid(anova_mod))

# Levene's test for homogeneity of variance
leveneTest(pain ~ trt, data = painstudy)
```

**Assumption Check Conclusion:** Looking at the Q-Q plot, the points follow the line reasonably well, and the Shapiro-Wilk test gives a p-value of 0.111, so normality seems fine. For equal variances, Levene's test gives p = 0.075, which is above 0.05, so we can assume the variances are roughly equal. Both assumptions are satisfied, so we can use the standard one-way ANOVA.

### ANOVA Results

```{r anova_results_3a, comment = NA, strip.white = TRUE}
summary(anova_mod)
```

**Interpretation:** The ANOVA gives an F-statistic of 98.92 with a p-value that is essentially zero (< 2e-16). This is well below our significance level of 0.05, so we reject the null hypothesis. There is a significant difference in mean pain scores between at least two of the treatments.

### Post-Hoc Analysis: Tukey's HSD

Since the ANOVA result is significant, we use Tukey's HSD test to see which specific pairs of treatments differ.

```{r tukey_3a, comment = NA, strip.white = TRUE}
tukey_result <- TukeyHSD(anova_mod)
tukey_result
```

```{r tukey_plot_3a, fig.width=5, fig.height=3.5, fig.align='center'}
plot(tukey_result, las = 1, cex.axis = 0.8)
```

**Post-Hoc Conclusion:** The Tukey test shows that all three pairwise comparisons are significant:

* **B vs A:** p = 0.0001, with the 95% CI [0.48, 1.76] above zero. Treatment B has higher pain scores than A.
* **C vs A:** p < 0.0001, with the 95% CI [-3.20, -1.92] below zero. Treatment C has lower pain scores than A.
* **C vs B:** p < 0.0001, with the 95% CI [-4.32, -3.04] below zero. Treatment C has lower pain scores than B.

**Overall Conclusion for 3A:** All three treatments are significantly different from each other. Treatment C is the best for reducing pain (mean = 3.64), Treatment A is in the middle (mean = 6.20), and Treatment B actually has the highest pain scores (mean = 7.32). So if the goal is to minimize pain, Treatment C would be the recommended choice.

## 3B: Randomized Block Design (RBD)

### Study Design Definitions

* **Research Question:** Is there a significant difference in shoe wear between material A and material B?
* **Experimental Units:** Each foot (left or right) of a boy.
* **Response Variable:** Shoe wear (`wearA`, `wearB`), a continuous measurement.
* **Treatment Factor:** Shoe material, with two levels: A and B.
* **Block Variable:** `boy` — each boy serves as a block.

### Why Blocking is Appropriate

Blocking by boy makes sense here because each boy has unique characteristics (weight, activity level, walking style, etc.) that could affect shoe wear. By having each boy wear both materials (one on each foot), we control for individual differences. This is a **within-subject design** where each subject acts as their own control, which typically reduces error variance and increases the power to detect a treatment effect.

### Data Preparation

The original data is in wide format (one row per boy). We need to reshape it to long format for proper ANOVA analysis.

```{r reshape_3b}
data("shoedat_obs", package = "scidesignR")

# Reshape to long format
shoe_long <- shoedat_obs %>%
  pivot_longer(
    cols = c(wearA, wearB),
    names_to = "material",
    values_to = "wear",
    names_prefix = "wear"
  ) %>%
  mutate(
    boy = factor(boy),
    material = factor(material)
  )

knitr::kable(head(shoe_long, 10), caption = "Reshaped Shoe Data (First 10 Rows)")
```

### Blocked Model: Response ~ Treatment + Block

```{r blocked_model_3b, comment = NA, strip.white = TRUE}
blocked_mod <- aov(wear ~ material + boy, data = shoe_long)
summary(blocked_mod)
```

**Interpretation:** In the blocked model, material has F = 1.74 with p = 0.220, which is not significant at $\alpha = 0.05$. The boy (block) effect has F = 0.43 with p = 0.888, indicating that the boys do not differ much from each other in terms of overall wear.

### Comparison with Unblocked Analysis

To see what blocking changes, we fit a model that ignores the blocking structure.

```{r unblocked_model_3b, comment = NA, strip.white = TRUE}
unblocked_mod <- aov(wear ~ material, data = shoe_long)
summary(unblocked_mod)
```

**Comparison:** In the unblocked model, material has F = 2.43 with p = 0.136, which is still not significant but has a lower p-value than the blocked model. This happens because the blocking factor (boy) does not explain much variation in this dataset — the boys are fairly similar in their overall wear levels. When blocking doesn't capture real between-block differences, it can actually increase the MSE by reducing the degrees of freedom for error without a corresponding reduction in error variance.

### Conclusion for 3B

Both models lead to the same conclusion: there is no statistically significant difference in wear between materials A and B. However, this dataset illustrates an important point — blocking is beneficial when there is substantial between-block variation to account for. In this case, the boys were relatively homogeneous, so blocking did not improve our analysis. In practice, blocking should be used when we expect meaningful differences between blocks (e.g., if the boys had very different activity levels or weights).

## 3C: Latin Square Design

### Study Design Definitions

* **Research Question:** Is there a significant effect of treatment on serum calcium levels in dogs?
* **Experimental Units:** Each dog-week combination (a single observation of one dog in one week).
* **Response Variable:** Serum calcium level (`value`), measured in mg-percent.
* **Treatment Factor:** `Treat`, with four levels: A, B, C, D.
* **Row Blocking Factor:** `Dogs` — blocks for individual dog differences (e.g., baseline calcium levels, metabolism).
* **Column Blocking Factor:** `Weeks` — blocks for time effects (e.g., seasonal variation, cumulative effects).

### Why Latin Square?

A Latin Square design is used when we need to control for two sources of nuisance variation simultaneously. Here, both individual dog differences and time (week) effects might influence serum calcium levels. By using a Latin Square, each treatment appears exactly once in each row (dog) and each column (week), so any dog or week effects are balanced out across treatments.

### Model Fitting

```{r latinsq_model_3c, comment = NA, strip.white = TRUE}
data("BR_LatSq", package = "scidesignR")

# Convert to factors
BR_LatSq$Dogs <- factor(BR_LatSq$Dogs)
BR_LatSq$Weeks <- factor(BR_LatSq$Weeks)
BR_LatSq$Treat <- factor(BR_LatSq$Treat)

# Fit Latin Square model
lat_sq_mod <- aov(value ~ Dogs + Weeks + Treat, data = BR_LatSq)
summary(lat_sq_mod)
```

### Interpretation

From the ANOVA table:

* **Dogs (rows):** F = 1.79, p = 0.248 — not significant. The dogs don't differ much in their overall calcium levels.
* **Weeks (columns):** F = 0.35, p = 0.794 — not significant. No meaningful time trend across the four weeks.
* **Treatment:** F = 3.01, p = 0.116 — not significant at $\alpha = 0.05$, but it has the largest F-value among the three factors.

```{r treat_means_3c}
# Treatment means
treat_means <- aggregate(value ~ Treat, data = BR_LatSq, FUN = mean)
knitr::kable(treat_means, col.names = c("Treatment", "Mean Serum Calcium"),
             caption = "Mean Serum Calcium by Treatment")
```

Looking at the treatment means, Treatments B and D have higher mean serum calcium (15.55) compared to A (14.75) and C (14.28). However, since the p-value is 0.116, we cannot conclude that these differences are statistically significant.

### Conclusion for 3C

Based on the Latin Square analysis, there is no statistically significant treatment effect on serum calcium levels (p = 0.116). While treatments B and D appear to produce higher calcium levels than A and C, the evidence is not strong enough to reject the null hypothesis at the 0.05 significance level. The blocking factors (dogs and weeks) also did not show significant effects, suggesting that the experimental units were fairly homogeneous across these dimensions.



# Part 4: Factorial Design at Two Levels
## Model Setup & Estimation
* **Factors:**
    * **Factor A:** Level of food diary
    * **Factor B:** Level of increasing activity
    * **Factor C:** Level of home visit

* **Levels:** Each factor has exactly two levels: a low level (-1) and a high level (1).

* **Response Variable ($y$):** Weight loss in kilograms (kg). Note that a higher positive value indicates greater weight loss, which is the desired outcome.

* This is a full \(2^3\) factorial design with three factors, \(A\), \(B\), and \(C\), each at two levels, giving \(2^3 = 8\) treatment combinations. The experimental unit is one study subject/experimental run under a particular combination of factor levels, and the response is weight loss in kilograms. No blocking variable is provided in `wtlossdat`, so the data are analyzed as an unblocked completely randomized factorial design. Since there is only one observation for each treatment combination, this is an unreplicated factorial experiment.

\vspace{-1.5em}
```{r part4_data}
data("wtlossdat", package = "scidesignR")

full_model <- lm(y ~ A * B * C, data = wtlossdat)

# Calculate effects (2 * coefficient)
effects <- 2 * coef(full_model)[-1]

knitr::kable(
  as.data.frame(effects),
  col.names = c("Effect Estimate"),
  caption = "Main Effects and Interactions Estimates"
)
```

In a two-level factorial design with factor levels coded as -1 and 1, the effect size is exactly twice the corresponding regression coefficient from a linear model.
**Interpretation:**

* **Factor B (Increasing activity):** Has a surprisingly strong negative main effect (-1.975), meaning that increasing the activity level actually reduced the amount of weight lost in this specific dataset.

* **Factor C (Home visit):** Has a positive main effect (0.575), meaning higher levels of home visits increase weight loss.

* **Interaction A:C**: There is a notable positive interaction between food diary and home visits.


Because this is a single unreplicated $2^3$ design, fitting a full model with the three-way interaction ($A \times B \times C$) leaves zero degrees of freedom for the residuals (error term), making it impossible to calculate p-values.

To resolve this, we will pool the highest-order interaction (the three-way interaction `A:B:C`) into the error term. This gives us 1 degree of freedom for the residuals, allowing us to evaluate the main effects and two-way interactions.

```{r ANOVA part}
# model without the 3-way interaction to allow 1 DF for error
mod_reduced <- aov(y ~ (A + B + C)^2, data = wtlossdat)
summary(mod_reduced)
```

**Interpretation:** Based on the ANOVA table, Factor B (Increasing activity) is marginally significant (p = 0.0563), falling just short of the standard 0.05 threshold. Factor C (Home visit) and the A:C interaction also show notable effect estimates but are not statistically significant (p = 0.1738). It is important to note that with only 1 degree of freedom for error, the statistical power of this test is severely limited, making it difficult to detect true effects at standard significance levels.

## Visualization and Recommendation

```{r Main effect, fig.height = 3.3, fig.width=5.5, fig.align='center'}
data("wtlossdat", package = "scidesignR")

wtlossdat_long <- wtlossdat %>%
  pivot_longer(
    cols = c(A, B, C),
    names_to = "Factor",
    values_to = "Level"
  ) %>%
  mutate(
    Factor = factor(
      Factor,
      levels = c("A", "B", "C"),
      labels = c("A: Food Diary", "B: Increasing Activity", "C: Home Visit")
    ),
    Level = factor(Level,
                   levels = c(-1, 1),
                   labels = c("Low (-1)", "High (1)"))
  ) %>%
  group_by(Factor, Level) %>%
  summarize(Mean_y = mean(y), .groups = "drop")

ggplot(wtlossdat_long, aes(x = Level, y = Mean_y, group = 1)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  facet_wrap(~Factor) +
  theme_bw() +
  labs(
    title = "Main Effects Plot",
    x = "Factor Level",
    y = "Mean Weight Loss"
  )

ac_plot_data <- wtlossdat %>%
  group_by(A, C) %>%
  summarize(Mean_y = mean(y), .groups = "drop") %>%
  mutate(
    A = factor(A, levels = c(-1, 1), labels = c("Low (-1)", "High (1)")),
    C = factor(C, levels = c(-1, 1), labels = c("Low (-1)", "High (1)"))
  )

```
**Main effects:** The main effects plot shows that B (Increasing Activity) has the strongest effect, and its high level reduces the response. In contrast, A (Food Diary) and C (Home Visit) both improve weight loss when set at their high levels.

```{r Interactions, fig.height = 3, fig.width=5, fig.align='center'}
ggplot(ac_plot_data, aes(x = A, y = Mean_y, group = C, linetype = C, shape = C)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  theme_bw() +
  labs(
    title = "Interaction Plot for A and C",
    x = "A: Food Diary",
    y = "Mean Weight Loss",
    linetype = "C: Home Visit",
    shape = "C: Home Visit"
  )
```

**Interaction:** The interaction plot suggests that **A and C** interact, since the lines are not parallel. This means the effect of the food diary depends on whether home visits are also used, with the best results occurring when both are at their high levels.

**Recommendation:** To improve the outcome, the recommended settings are **A = high, B = low, and C = high**. These settings give the largest observed weight loss in the data, with ($y = 2.6$).


\textbf{References}


\newpage
\appendix

# Appendix: R Code

```{r appendix_code, ref.label = setdiff(knitr::all_labels(), "appendix_code"), echo = TRUE, eval = FALSE}
```
