---
title: 'STA305H5S: Project'
author: "Jack Chamoun Barbar, Mir Asim Ali, Parth Tyagi"
date: "April 5, 2026"
output:
  pdf_document:
    number_sections: true
  word_document: default
spacing: double
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- "\\fancyhead[L]{STA305H5-Experimental Design}"
- \fancyfoot[CO,CE]{}
- \fancyfoot[LE,RO]{\thepage}
- \usepackage{setspace}\doublespacing
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

# The libraries needed for our project
library(scidesignR)
library(ggplot2)
library(dplyr)
library(tidyr)
library(car)
library(coin)
```

\newpage 
\tableofcontents 
\newpage

# Part 1: Comparing Two Treatments

## Study Design Definitions

* **Research Question:** Is there a statistically significant difference in the mean wheat yield between Treatment A and Treatment B? In other words, we are testing: $H_0:\mu_A=\mu_B$ vs $H_1:\mu_A \neq \mu_B$, where
$\mu_A$ and $\mu_B$ refer to the mean wheat yields of treatments A and B, respectively.
* **Experimental Units:** The wheat plots, which are randomly assigned to receive the treatments.
* **Response Variable:** Wheat yield. It is represented as `fert` in the `fertdat`
dataset. Note that wheat yield is a quantitative continuous response variable
(as the data are collected by measuring and expressed on a continuous scale).
* **Treatment Factor:** Level/type of fertilizer, which is expressed as `trt` the
`fertdat` dataset, with two levels (A and B).

## Exploratory Data Analysis (EDA)
Exploratory Data Analysis (EDA) is the initial phase of data science that uses descriptive statistics and visual techniques  to uncover patterns, identify anomalies, and understand the underlying structure of a dataset. It is essential because it prevents flawed modeling by informing data cleaning processes, testing necessary statistical assumptions, and guiding the selection of appropriate statistical tests.

\large
```{r eda, fig.width=4, fig.height=3.5, fig.align='center'}
# Load the dataset
data("fertdat", package = "scidesignR")


summary_stats <- fertdat %>%
  group_by(trt) %>%
  summarize(
    n = n(),
    Mean = round(mean(fert, na.rm = TRUE), 4),
    SD = round(sd(fert, na.rm = TRUE), 4),
    Min = min(fert, na.rm = TRUE),
    Max = max(fert, na.rm = TRUE)
  )
knitr::kable(summary_stats, caption = "Summary Statistics of Wheat Yield by Fertilizer")

# A boxplot
ggplot(fertdat, aes(x = trt, y = fert, fill = trt)) +
  geom_boxplot(alpha = 0.7) +
  theme_minimal() +
  labs(title = "Wheat Yield Distribution by Fertilizer Type",
       x = "Type of Fertilizer",
       y = "Wheat Yield") +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5))
```
\normalsize

## Checking Assumptions

To proceed with a standard two-sample t-test, we must check for **normality** of the residuals
and **homoscedasticity** of variance, i.e., constant variance.


```{r model_setup, include=FALSE}
mod <- lm(fert ~ trt, data = fertdat)
```

```{r qq_plot, fig.width=4.5, fig.height=3.8, fig.align='center', out.width='65%'}
qqnorm(resid(mod), main = "Q-Q Plot of Residuals"); qqline(resid(mod), col = "red")
```
\normalsize

\newpage

```{r residual_plot, fig.width=4.5, fig.height=4, fig.align='center'}
plot(fitted(mod), resid(mod),
     main = "Residuals vs Fitted",
     xlab = "Fitted Values",
     ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)
```
Since plots can be subjective, it is better to supplement with more formal/robust tests. Now, we use the Shapiro and Levene's tests, which assess for normality and equality of population variances
(i.e, $\sigma^2_A = \sigma^2_B$), respectively. We chose Levene's test as it is very robust against
non-normality, unlike Bartlett's test.

\vspace{-2em}
```{r Formal tests, comment = NA, strip.white = TRUE}
# Normality check
shapiro.test(resid(mod))

# Levene's test for equal variances
leveneTest(fert ~ trt, data = fertdat)
```

**Assumption Check Conclusion:** The assumption of normality is met, as the Q-Q plot points closely follow the theoretical line and the Shapiro-Wilk test yields a non-significant p-value of 0.8418. However, the assumption of homoscedasticity is violated because the Residuals vs Fitted plot displays an unequal spread and Levene’s test results in a significant p-value of 0.04837. Thus, we should proceed with a Welch's two-sample t-test instead of a standard two-sample t-test to account for the unequal variances ($\sigma^2_A \neq \sigma^2_B$).

## Hypothesis Testing

Let $\mu_A$ and $\mu_B$ be the true mean wheat yields for treatments A and B, respectively.

* **Null Hypothesis ($H_0$):** $\mu_A = \mu_B$
* **Alternative Hypothesis ($H_1$):** $\mu_A \neq \mu_B$

\vspace{-2.5em}
```{r Testing, comment = NA, strip.white = TRUE}
# Welch's two-sample t-test
welch_test <- t.test(fert ~ trt, data = fertdat, var.equal = FALSE)
welch_test
```

## Results and Interpretation

Based on the R output for Welch's two-sample t-test:

* **Test Statistic:** $t = `r round(welch_test$statistic, 4)`$
* **Degrees of Freedom:** $df = `r round(welch_test$parameter, 4)`$
* **P-value:** $`r format.pval(welch_test$p.value)`$
* **95% Confidence Interval:** $[`r round(welch_test$conf.int[1], 4)`, `r round(welch_test$conf.int[2], 4)`]$

**Conclusion:** With a large p-value of 0.6168, we fail to reject the null hypothesis, as there is insufficient evidence to conclude a significant difference in true mean wheat yields between fertilizers A and B. Furthermore, we are 95% confident that the true difference in mean yields ($\mu_A - \mu_B$) lies between -5.86 and 9.26, an interval that includes zero and is consistent with the lack of a statistically significant difference.


# Part 2: Power & Sample Size

## Practical Effect Size and Required Sample Size
We use the built-in `power.t.test()` function in R to compute the Effect Size
and the sample size required to achieve 80% power.

\vspace{-3em}
```{r effect-size, comment = NA, strip.white = TRUE}
delta_est <- 22.1833 - 20.4833

sd_pooled <- sigma(mod)

alpha_level <- 0.05; target_power <- 0.80

exact_calc <- power.t.test(delta = delta_est,
                           sd = sd_pooled,
                           sig.level = alpha_level,
                           power = target_power,
                           type = "two.sample",
                           alternative = "two.sided")

exact_calc
```

Based on the linear model from Part 1, the estimated (unstandardized) effect size ($\Delta = |\bar{x}_A - \bar{x}_B|$) is `r round(delta_est, 4)`, and the pooled standard deviation is $s_{pooled}\approx$ `r round(sd_pooled, 4)`.
Thus, our standardized effect size (Cohen's $d$) is $d = \frac{|\bar{x}_A - \bar{x}_B|}{s_{pooled}} = \frac{1.7}{5.6508} \approx 0.3008$. The required sample size to achieve 80% power at a significance level of $\alpha = 0.05$ is **175** (after rounding up to the next integer) **per group**.

## Simulated Power
This code calculates statistical power by simulating two data groups 10,000 times, running a t-test on each iteration, and finding the proportion of p-values that fall below the significance threshold $\alpha=0.05$.

```{r Simulated Power}
set.seed(123)  # So that whoever runs this gets same exact number
m <- 10000
p_values <- numeric(m) # To store the p-values in this (will use after the loop)

for(i in 1:m) {
  sim_A <- rnorm(175, mean = 0, sd = sd_pooled)
  sim_B <- rnorm(175, mean = delta_est, sd = sd_pooled)
  test_result <- t.test(sim_A, sim_B, var.equal = FALSE)
  p_values[i] <- test_result$p.value
}

simulated_power <- mean(p_values < alpha_level) # Proportion of times we reject
cat("Simulated Power:", simulated_power, "\n")
```

## Discussion on Changing Parameters

* **If the effect size is smaller:** The required sample size **increases**. A smaller true difference is harder to detect against random noise, so more data is needed to correctly reject the null hypothesis.
* **If the variability is larger:** The required sample size **increases**. A higher standard deviation creates more overlap between the two groups, requiring more observations to clearly isolate the treatment effect.

# Part 3: More Than Two Treatments

## 3A — One-Way ANOVA

### Study Design

* **Research Question:** Is there a statistically significant difference in mean pain scores across the treatment groups? Formally: $H_0: \mu_1 = \mu_2 = \mu_3 = \mu_4$ vs $H_1:$ at least one $\mu_i$ differs.
* **Experimental Units:** Individual patients assigned to each treatment.
* **Response Variable:** Pain score (`pain`), measured on a scale of 0 to 10.
* **Treatment Factor:** Treatment group (`trt`), a categorical factor with four levels.

### Exploratory Data Analysis

```{r part3a_eda, fig.width=5, fig.height=3.5, fig.align='center'}
data("painstudy", package = "scidesignR")

# Summary statistics by treatment group
pain_summary <- painstudy %>%
  group_by(trt) %>%
  summarize(
    n = n(),
    Mean = round(mean(pain, na.rm = TRUE), 4),
    SD = round(sd(pain, na.rm = TRUE), 4),
    Min = min(pain, na.rm = TRUE),
    Max = max(pain, na.rm = TRUE)
  )
knitr::kable(pain_summary, caption = "Summary Statistics of Pain Scores by Treatment")

# Boxplot
ggplot(painstudy, aes(x = factor(trt), y = pain, fill = factor(trt))) +
  geom_boxplot(alpha = 0.7) +
  theme_minimal() +
  labs(title = "Pain Score Distribution by Treatment Group",
       x = "Treatment Group",
       y = "Pain Score") +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5))
```

### Checking Assumptions

To use a one-way ANOVA, we need to verify **normality** of the residuals and **homogeneity of variances** across groups.

```{r part3a_assumptions, fig.width=4.5, fig.height=3.8, fig.align='center', out.width='65%'}
mod_pain <- lm(pain ~ factor(trt), data = painstudy)

# Q-Q plot of residuals
qqnorm(resid(mod_pain), main = "Q-Q Plot of Residuals (Pain Study)")
qqline(resid(mod_pain), col = "red")
```

```{r part3a_resid_plot, fig.width=4.5, fig.height=3.8, fig.align='center'}
plot(fitted(mod_pain), resid(mod_pain),
     main = "Residuals vs Fitted (Pain Study)",
     xlab = "Fitted Values",
     ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)
```

```{r part3a_formal_tests, comment = NA, strip.white = TRUE}
# Shapiro-Wilk test for normality of residuals
shapiro.test(resid(mod_pain))

# Levene's test for homogeneity of variances
leveneTest(pain ~ factor(trt), data = painstudy)
```

**Assumption Check Conclusion:** If the Shapiro-Wilk test yields a non-significant p-value and the Q-Q plot shows points closely following the theoretical line, then normality is reasonable. If Levene's test is non-significant, the equal variance assumption holds, and we may proceed with a standard one-way ANOVA. If either assumption is violated, we note it and consider robust alternatives (e.g., Welch's ANOVA).

### One-Way ANOVA

```{r part3a_anova, comment = NA, strip.white = TRUE}
aov_pain <- aov(pain ~ factor(trt), data = painstudy)
summary(aov_pain)
```

**Interpretation:** The ANOVA F-test evaluates whether the mean pain scores differ significantly across the treatment groups. If the p-value is below 0.05, we reject the null hypothesis and conclude that at least one treatment group has a significantly different mean pain score. Otherwise, we fail to reject $H_0$.

### Post-Hoc Comparisons

If the ANOVA result is significant, we perform Tukey's Honest Significant Difference (HSD) test to identify which specific pairs of treatments differ.

```{r part3a_tukey, comment = NA, strip.white = TRUE}
tukey_result <- TukeyHSD(aov_pain)
tukey_result
```

```{r part3a_tukey_plot, fig.width=5.5, fig.height=4, fig.align='center'}
plot(tukey_result, las = 1, cex.axis = 0.7)
```

**Post-Hoc Interpretation:** Each pairwise comparison includes a 95% confidence interval for the difference in means and an adjusted p-value. Pairs whose confidence intervals do not contain zero indicate a statistically significant difference. The Tukey plot visually displays these intervals, with any interval not crossing zero highlighting a significant pairwise contrast.

\newpage

## 3B — Randomized Block Design (RBD)

### Study Design

* **Research Question:** Is there a significant difference in wear between shoe materials A and B, after controlling for boy-to-boy variability?
* **Experimental Units:** Each foot (left or right) of a boy.
* **Response Variable:** Wear of the shoe material (measured as `wearA` and `wearB`).
* **Treatment Factor:** Shoe material (A vs B).
* **Block Variable:** Boy (`boy`). Each boy wears both materials (one on each foot), so each boy serves as a block. This is appropriate because individual boys may have different activity levels, walking habits, or body weights that affect wear, and blocking eliminates this source of variability.

### Data Preparation

The dataset `shoedat_obs` stores the two measurements side by side. We reshape it into long format for modeling.

```{r part3b_data}
data("shoedat_obs", package = "scidesignR")

# Reshape to long format
shoe_long <- shoedat_obs %>%
  pivot_longer(
    cols = c(wearA, wearB),
    names_to = "material",
    values_to = "wear",
    names_prefix = "wear"
  )
shoe_long$boy <- factor(shoe_long$boy)
shoe_long$material <- factor(shoe_long$material)

knitr::kable(
  shoe_long %>%
    group_by(material) %>%
    summarize(
      n = n(),
      Mean = round(mean(wear), 4),
      SD = round(sd(wear), 4)
    ),
  caption = "Summary Statistics of Wear by Material"
)
```

### Why Blocking is Appropriate

Each boy acts as his own control: he wears material A on one foot and material B on the other. Boys differ in weight, stride, and activity level, all of which affect shoe wear. By blocking on boy, we remove this between-subject variability from the error term, making the comparison of materials A vs B more precise. Think of it like comparing two paint colors on the same wall rather than on different walls — the wall-to-wall differences (lighting, texture) are eliminated.

### Fitting the RBD Model

```{r part3b_rbd, comment = NA, strip.white = TRUE}
mod_rbd <- lm(wear ~ material + boy, data = shoe_long)
anova(mod_rbd)
```

**Interpretation:** The ANOVA table tests the treatment effect (material) after accounting for block effects (boy). A significant p-value for `material` indicates that the two shoe materials produce different amounts of wear when boy-to-boy variability is controlled. The block effect (boy) is expected to be significant, confirming that blocking was effective.

### Comparison with Unblocked Analysis

```{r part3b_unblocked, comment = NA, strip.white = TRUE}
mod_unblocked <- lm(wear ~ material, data = shoe_long)
anova(mod_unblocked)
```

**Blocked vs Unblocked:** In the unblocked analysis, all variability between boys is pooled into the residual error term, inflating the MSE and making it harder to detect a real treatment effect. The blocked model separates block variability from error, yielding a smaller residual MSE and therefore a more powerful F-test. This illustrates the key benefit of blocking: if the blocking factor explains substantial variability, blocking increases the precision of treatment comparisons. In simple terms, blocking is like noise-canceling headphones — it removes the background noise (boy differences) so you can hear the signal (material difference) more clearly.

\newpage

## 3C — Latin Square Design

### Study Design

* **Research Question:** Is there a significant difference in mean serum calcium levels across the treatments, after controlling for both dog and week effects?
* **Experimental Units:** A dog-week combination (i.e., one dog during one specific week).
* **Response Variable:** Serum calcium level in mg-per cent (`value`).
* **Row Blocking Factor:** Dog (`Dogs`) — different dogs may have different baseline calcium levels.
* **Column Blocking Factor:** Week (`Weeks`) — calcium levels may drift over time.
* **Treatment Factor:** Treatment (`Treat`) — the intervention applied.

The Latin Square design simultaneously blocks on two nuisance factors (dogs and weeks) using each treatment exactly once in each row and each column. This is like seating guests at a dinner table so that every guest sits next to every other guest exactly once — each blocking factor is perfectly balanced with treatments.

### Exploratory Data Analysis

```{r part3c_eda, fig.width=5, fig.height=3.5, fig.align='center'}
data("BR_LatSq", package = "scidesignR")

BR_LatSq$Dogs <- factor(BR_LatSq$Dogs)
BR_LatSq$Weeks <- factor(BR_LatSq$Weeks)
BR_LatSq$Treat <- factor(BR_LatSq$Treat)

ls_summary <- BR_LatSq %>%
  group_by(Treat) %>%
  summarize(
    n = n(),
    Mean = round(mean(value, na.rm = TRUE), 4),
    SD = round(sd(value, na.rm = TRUE), 4)
  )
knitr::kable(ls_summary, caption = "Summary Statistics of Serum Calcium by Treatment")

ggplot(BR_LatSq, aes(x = Treat, y = value, fill = Treat)) +
  geom_boxplot(alpha = 0.7) +
  theme_minimal() +
  labs(title = "Serum Calcium by Treatment (Latin Square)",
       x = "Treatment",
       y = "Serum Calcium (mg%)") +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5))
```

### Latin Square Model

```{r part3c_model, comment = NA, strip.white = TRUE}
mod_ls <- lm(value ~ Dogs + Weeks + Treat, data = BR_LatSq)
anova(mod_ls)
```

**Interpretation:** The ANOVA table tests the treatment effect after adjusting for both row (Dogs) and column (Weeks) blocking factors. A significant p-value for `Treat` indicates that at least one treatment produces a different mean serum calcium level than the others, after accounting for dog-to-dog and week-to-week variability. Significant effects for `Dogs` or `Weeks` confirm that blocking on those factors was effective in reducing unexplained variation.



# Part 4: Factorial Design at Two Levels
## Model Setup & Estimation
* **Factors:**
    * **Factor A:** Level of food diary
    * **Factor B:** Level of increasing activity
    * **Factor C:** Level of home visit

* **Levels:** Each factor has exactly two levels: a low level (-1) and a high level (1).

* **Response Variable ($y$):** Weight loss in kilograms (kg). Note that a higher positive value indicates greater weight loss, which is the desired outcome.

* This is a full \(2^3\) factorial design with three factors, \(A\), \(B\), and \(C\), each at two levels, giving \(2^3 = 8\) treatment combinations. The experimental unit is one study subject/experimental run under a particular combination of factor levels, and the response is weight loss in kilograms. No blocking variable is provided in `wtlossdat`, so the data are analyzed as an unblocked completely randomized factorial design. Since there is only one observation for each treatment combination, this is an unreplicated factorial experiment.

\vspace{-1.5em}
```{r part4_data}
data("wtlossdat", package = "scidesignR")

full_model <- lm(y ~ A * B * C, data = wtlossdat)

# Calculate effects (2 * coefficient)
effects <- 2 * coef(full_model)[-1]

knitr::kable(
  as.data.frame(effects),
  col.names = c("Effect Estimate"),
  caption = "Main Effects and Interactions Estimates"
)
```

In a two-level factorial design with factor levels coded as -1 and 1, the effect size is exactly twice the corresponding regression coefficient from a linear model.
**Interpretation:**

* **Factor B (Increasing activity):** Has a surprisingly strong negative main effect (-1.975), meaning that increasing the activity level actually reduced the amount of weight lost in this specific dataset.

* **Factor C (Home visit):** Has a positive main effect (0.575), meaning higher levels of home visits increase weight loss.

* **Interaction A:C**: There is a notable positive interaction between food diary and home visits.


Because this is a single unreplicated $2^3$ design, fitting a full model with the three-way interaction ($A \times B \times C$) leaves zero degrees of freedom for the residuals (error term), making it impossible to calculate p-values.

To resolve this, we will pool the highest-order interaction (the three-way interaction `A:B:C`) into the error term. This gives us 1 degree of freedom for the residuals, allowing us to evaluate the main effects and two-way interactions.

```{r ANOVA part}
# model without the 3-way interaction to allow 1 DF for error
mod_reduced <- aov(y ~ (A + B + C)^2, data = wtlossdat)
summary(mod_reduced)
```

**Interpretation:** Based on the ANOVA table, Factor B (Increasing activity) is marginally significant (p = 0.0563), falling just short of the standard 0.05 threshold. Factor C (Home visit) and the A:C interaction also show notable effect estimates but are not statistically significant (p = 0.1738). It is important to note that with only 1 degree of freedom for error, the statistical power of this test is severely limited, making it difficult to detect true effects at standard significance levels.

## Visualization and Recommendation

```{r Main effect, fig.height = 3.3, fig.width=5.5, fig.align='center'}
data("wtlossdat", package = "scidesignR")

wtlossdat_long <- wtlossdat %>%
  pivot_longer(
    cols = c(A, B, C),
    names_to = "Factor",
    values_to = "Level"
  ) %>%
  mutate(
    Factor = factor(
      Factor,
      levels = c("A", "B", "C"),
      labels = c("A: Food Diary", "B: Increasing Activity", "C: Home Visit")
    ),
    Level = factor(Level,
                   levels = c(-1, 1),
                   labels = c("Low (-1)", "High (1)"))
  ) %>%
  group_by(Factor, Level) %>%
  summarize(Mean_y = mean(y), .groups = "drop")

ggplot(wtlossdat_long, aes(x = Level, y = Mean_y, group = 1)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  facet_wrap(~Factor) +
  theme_bw() +
  labs(
    title = "Main Effects Plot",
    x = "Factor Level",
    y = "Mean Weight Loss"
  )

ac_plot_data <- wtlossdat %>%
  group_by(A, C) %>%
  summarize(Mean_y = mean(y), .groups = "drop") %>%
  mutate(
    A = factor(A, levels = c(-1, 1), labels = c("Low (-1)", "High (1)")),
    C = factor(C, levels = c(-1, 1), labels = c("Low (-1)", "High (1)"))
  )

```
**Main effects:** The main effects plot shows that B (Increasing Activity) has the strongest effect, and its high level reduces the response. In contrast, A (Food Diary) and C (Home Visit) both improve weight loss when set at their high levels.

```{r Interactions, fig.height = 3, fig.width=5, fig.align='center'}
ggplot(ac_plot_data, aes(x = A, y = Mean_y, group = C, linetype = C, shape = C)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  theme_bw() +
  labs(
    title = "Interaction Plot for A and C",
    x = "A: Food Diary",
    y = "Mean Weight Loss",
    linetype = "C: Home Visit",
    shape = "C: Home Visit"
  )
```

**Interaction:** The interaction plot suggests that **A and C** interact, since the lines are not parallel. This means the effect of the food diary depends on whether home visits are also used, with the best results occurring when both are at their high levels.

**Recommendation:** To improve the outcome, the recommended settings are **A = high, B = low, and C = high**. These settings give the largest observed weight loss in the data, with ($y = 2.6$).


\textbf{References}


\newpage
\appendix

# Appendix: R Code

```{r appendix_code, ref.label = setdiff(knitr::all_labels(), "appendix_code"), echo = TRUE, eval = FALSE}
```
